---

---

Downloads and cleans Colorado construction data from two federal sources with no
API yet, and loads it into the SDO Postgres database. Re-render this notebook
whenever Census/BPS publish a new annual vintage -- everything below is written
to run unattended top to bottom.

```{r}
#| label: setup

library(tidyverse)
library(readr)
library(readxl)
library(DBI)
library(sdotools) # sdo_db_connect()

raw_dir <- "raw_files"
dir.create(raw_dir, showWarnings = FALSE, recursive = TRUE)

download_once <- function(url, dest) {
  if (!file.exists(dest)) {
    ok <- tryCatch(
      {
        utils::download.file(url, dest, mode = "wb", quiet = TRUE)
        TRUE
      },
      error = function(e) FALSE
    )
    if (!ok) {
      if (file.exists(dest)) {
        file.remove(dest)
      }
      return(NA_character_)
    }
  }
  dest
}
```

## Census C30: nonresidential value put in place (state only)

No county breakdown exists in C30 at all -- it's not published below the state
level. The Census site splits history into non-overlapping vintage files that
get re-cut over time; three are stitched together here for 1993-present. The
row label ("Colorado") is matched by text, not column position -- the state
name lands in a different column across the three vintages' layouts.

```{r}
#| label: c30_download_parse

read_nrstate_co <- function(path) {
  raw <- read_excel(path, sheet = 1, col_names = FALSE)
  year_hdr <- raw[4, ]
  year_cols <- which(!is.na(suppressWarnings(as.numeric(unlist(year_hdr)))))
  years <- as.integer(unlist(year_hdr[year_cols]))

  name_col <- which(vapply(
    raw,
    \(col) any(col == "Colorado", na.rm = TRUE),
    logical(1)
  ))
  stopifnot(length(name_col) == 1)
  co_row <- raw[which(raw[[name_col]] == "Colorado"), ]
  stopifnot(nrow(co_row) == 1)

  tibble(
    year = years,
    value_millions = as.numeric(unlist(co_row[1, year_cols]))
  )
}

c30_sources <- tribble(
  ~url                                                           , ~file                       ,
  "https://www.census.gov/construction/c30/xls/nrstatehs.xls"    , "nrstatehs_1993_2007.xls"   ,
  "https://www.census.gov/construction/c30/xlsx/nrstatehs1.xlsx" , "nrstatehs1_2008_2021.xlsx" ,
  "https://www.census.gov/construction/c30/xlsx/nrstate.xlsx"    , "nrstate_current.xlsx"
)

c30_paths <- map2_chr(c30_sources$url, c30_sources$file, \(u, f) {
  download_once(u, file.path(raw_dir, f))
})
if (anyNA(c30_paths)) {
  message(
    "C30 download failed: ",
    paste(c30_sources$file[is.na(c30_paths)], collapse = ", ")
  )
}

const_value_state <- c30_paths[!is.na(c30_paths)] |>
  map(read_nrstate_co) |>
  bind_rows() |>
  arrange(desc(year)) |>
  distinct(year, .keep_all = TRUE) |> # newer vintage wins if a future re-cut overlaps years
  arrange(year) |>
  mutate(
    state_fips = "08",
    value = value_millions * 1e6, # native unit is millions; stored here as actual dollars
    source = "census_c30_nrstate",
    category = "nres"
  ) |>
  select(year, state_fips, category, value, source)

message(sprintf(
  "C30 nonresidential value, CO: %d years (%d-%d)",
  nrow(const_value_state),
  min(const_value_state$year),
  max(const_value_state$year)
))
```

## Census/HUD BPS: residential building permits (state and county)

Annual files are stable-format ASCII, comma-delimited, 3 header lines then
data (spot-checked 1980/1991-2024: 29 state / 30 county columns throughout,
despite the CO county row count itself growing 62->64 around 2010 when
Broomfield County split off -- that's real, not a parsing artifact). Columns
6:17 (state) / 7:18 (county) are "Estimates with Imputation" --
Buildings/Units/Valuation x {1-unit, 2-units, 3-4 units, 5+ units} -- the
headline published series; "Reported Only" columns are dropped.

**Unit gotcha, verified empirically (not just from the Census PDFs, which
document both identically as "thousands of dollars"):** BPS *state* valuation
fields are thousands of dollars; BPS *county* valuation fields are already
actual dollars. Both are normalized to actual dollars below. The
state-vs-sum-of-counties reconciliation check further down would be off by a
factor of ~1000 if this were backwards -- it isn't, so treat that check as
the real proof, not the doc text.

```{r}
#| label: bps_download_parse

this_year <- as.integer(format(Sys.Date(), "%Y")) - 1

fetch_bps_annual <- function(geo, years) {
  prefix <- if (geo == "state") "st" else "co"
  base_url <- sprintf(
    "https://www2.census.gov/econ/bps/%s/",
    if (geo == "state") "State" else "County"
  )
  map(years, \(y) {
    fname <- sprintf("%s%da.txt", prefix, y)
    dest <- download_once(paste0(base_url, fname), file.path(raw_dir, fname))
    if (is.na(dest)) {
      return(NULL)
    }
    n_id_cols <- if (geo == "state") 5L else 6L
    # FIPS columns look numeric to read_csv's type guesser, which would drop
    # the zero-padding (state "08" -> 8, county "001" -> 1) -- force character.
    df <- suppressWarnings(read_csv(
      dest,
      skip = 3,
      col_names = FALSE,
      show_col_types = FALSE,
      col_types = cols(X2 = "c", .default = "c")
    ))
    # which() (not a logical mask) -- an NA in a logical subscript produces an
    # all-NA row rather than dropping it, and non-numeric leftover rows (blank
    # dividers, etc.) make as.integer() coercion produce exactly that NA.
    fips_num <- suppressWarnings(as.integer(df[[2]]))
    df <- df[which(fips_num == 8L), , drop = FALSE]
    if (nrow(df) == 0) {
      return(NULL)
    }
    tibble(
      year = y,
      state_fips = sprintf("%02d", as.integer(df[[2]])),
      county_fips = if (geo == "county") {
        sprintf("%03d", as.integer(df[[3]]))
      } else {
        NA_character_
      },
      units_1unit = as.numeric(df[[n_id_cols + 2]]),
      value_1unit = as.numeric(df[[n_id_cols + 3]]),
      units_2unit = as.numeric(df[[n_id_cols + 5]]),
      value_2unit = as.numeric(df[[n_id_cols + 6]]),
      units_3to4unit = as.numeric(df[[n_id_cols + 8]]),
      value_3to4unit = as.numeric(df[[n_id_cols + 9]]),
      units_5punit = as.numeric(df[[n_id_cols + 11]]),
      value_5punit = as.numeric(df[[n_id_cols + 12]])
    )
  }) |>
    compact() |>
    bind_rows()
}

bps_state_raw <- fetch_bps_annual("state", 1980:this_year)
bps_county_raw <- fetch_bps_annual("county", 1991:this_year)

tidy_bps <- function(df, value_scale) {
  df |>
    mutate(across(starts_with("value_"), \(v) v * value_scale)) |>
    mutate(
      totperm = units_1unit + units_2unit + units_3to4unit + units_5punit,
      totval = value_1unit + value_2unit + value_3to4unit + value_5punit
    )
}

const_permits_state <- tidy_bps(bps_state_raw, value_scale = 1000) |>
  mutate(source = "census_bps_state") |>
  select(year, state_fips, totperm, totval, units_1unit:value_5punit, source) |>
  arrange(year)

const_permits_county <- tidy_bps(bps_county_raw, value_scale = 1) |>
  mutate(source = "census_bps_county") |>
  select(
    year,
    state_fips,
    county_fips,
    totperm,
    totval,
    units_1unit:value_5punit,
    source
  ) |>
  arrange(year, county_fips)

message(sprintf(
  "BPS state permits, CO: %d years (%d-%d)",
  nrow(const_permits_state),
  min(const_permits_state$year),
  max(const_permits_state$year)
))
message(sprintf(
  "BPS county permits, CO: %d rows, %d counties, %d-%d",
  nrow(const_permits_county),
  n_distinct(const_permits_county$county_fips),
  min(const_permits_county$year),
  max(const_permits_county$year)
))
```

## Sanity check: state total vs. sum of counties

Same year, same (now-normalized) units -- these should track closely. County
reporting is opt-in, so a small gap is expected in thin years; anything large
signals a unit or parsing regression.

```{r}
#| label: reconcile

state_vs_county <- const_permits_county |>
  summarize(totval_county_sum = sum(totval, na.rm = TRUE), .by = year) |>
  inner_join(
    const_permits_state |> select(year, totval_state = totval),
    by = "year"
  ) |>
  mutate(pct_diff = (totval_county_sum - totval_state) / totval_state)

state_vs_county |> arrange(year)
```

## Write to Postgres

Full overwrite each run (small tables, annual data does get revised) --
matches `process_lehd`'s `econ.commute` pattern, not the newer
`sdo_db_write_table()` model-tracking wrapper (`econ.lehd_mjh` /
`econ.commute` have no `model_id` column, and downstream readers in
`state_forecast_2025` don't pass `filter_model_id`).

```{r}
#| label: write_postgres

pg <- sdo_db_connect()

dbWriteTable(
  pg,
  name = Id(schema = "econ", table = "construction_value_state"),
  const_value_state,
  overwrite = TRUE,
  row.names = FALSE
)

dbWriteTable(
  pg,
  name = Id(schema = "econ", table = "construction_permits_state"),
  const_permits_state,
  overwrite = TRUE,
  row.names = FALSE
)

dbWriteTable(
  pg,
  name = Id(schema = "econ", table = "construction_permits_county"),
  const_permits_county,
  overwrite = TRUE,
  row.names = FALSE
)

dbDisconnect(pg)
message("Wrote 3 tables to econ schema.")
```

## Local CSV mirror

Committed to git as an audit trail -- same pattern as `process_lehd`'s
`lehd_mjh.csv`.

```{r}
#| label: write_csv

write_csv(const_value_state, "construction_value_state.csv")
write_csv(const_permits_state, "construction_permits_state.csv")
write_csv(const_permits_county, "construction_permits_county.csv")
```
